Counts of Crimes and Their Subtypes in Toronto from 2014 to 2023

### 2024-09-26 paper/paper.qmd
Fourth Commit. Discussion, Intro, and Abstract to be finished

The remainder of this paper is structured as follows. @sec-data....

@@

The dataset being used here is downloaded from Toronto Open Data website via @TorontoOpenData. This dataset contains all the reported crimes happened in the city of Toronto from 2014 to 2023. This dataset is grouped by the year of the reported crime, its category and belonging subtype, as well as the count of the subtype being reported and cleared for that year, for each division. Because I am examining the crime pattern in the whole city, I have dropped the division information, and aggregated the existing data according to their subtype, and the year of the crime being reported. In the following sub-sections, I will go through all the variables used in this report, and provide some basic descriptive statistics about them. The first 5 rows of the cleaned data used for analysis is attached below.

```{python}
import pandas as pd
data = pd.read_csv('../data/analysis_data/analysis_data.csv')
data.head()
```

## Report_Year
Report Year variable is the year of crimes being reported. In this dataset, the data spans from 2014 to 2023, encompassing a ten-year period. No month or date information was given, thus there are only 10 different values for this variable, in a chronological order.

## Category
Category includes information about the nature of the crime. In total there are six crime categroies, they are Crimes Against Property, Crimes Against the Person, Other Federal Statute Violations, Other Criminal Code Violations, Controlled Drugs and Substances Act, and Criminal Code Traffic. They are listed in the table below.

```{python}
import pandas as pd
import plotly.graph_objects as go
data = pd.read_csv('../data/analysis_data/analysis_data.csv')
unique_category = data.CATEGORY.unique()

category_table = go.Figure(data=[go.Table(header=dict(values=['Categories']), 
                 cells=dict(values=[unique_category])) 
                     ]) 
category_table.show() 
```

## Subtype
Under each crime category, there exist multiple subtypes within each category. The following is an exhaustive table of all of the subtypes of crimes and their respective category.

```{python}
import pandas as pd
import plotly.graph_objects as go
data = pd.read_csv('../data/analysis_data/analysis_data.csv')

unique_subtypes_data = data.drop_duplicates(subset='SUBTYPE')
unique_subtypes_list = unique_subtypes_data['SUBTYPE'].tolist()
categories_list_for_unique_subtypes = unique_subtypes_data['CATEGORY'].tolist()

subtype_table = go.Figure(data=[go.Table(header=dict(values=['Categories','Subtypes']), 
                 cells=dict(values=[categories_list_for_unique_subtypes, unique_subtypes_list])) 
                     ]) 
subtype_table.show() 
#print(unique_subtypes_list)
#print(categories_list_for_unique_subtypes)
```

@@

# Results

>>>>

The remainder of this paper is structured as follows.

@@

The dataset being used here is downloaded from Toronto Open Data website via @TorontoOpenData. This dataset contains all the reported crimes happened in the city of Toronto from 2014 to 2023. This dataset is grouped by the year of the reported crime, its category and belonging subtype, as well as the count of the subtype being reported and cleared for that year, for each division. Because I am examining the crime pattern in the whole city, I have dropped the division information, and aggregated the existing data according to their subtype, and the year of the crime being reported. In the following sub-sections, I will go through all the variables used in this report, and provide some basic descriptive statistics about them. The first 5 rows of the cleaned data used for analysis is attached (@tbl-table1).

```{python}
#| label: tbl-table1
#| tbl-cap: Example of Cleaned Data
#| warning: false
#| echo: false

import pandas as pd
data = pd.read_csv('../data/analysis_data/analysis_data.csv', index_col=False)
data.index = data.index + 1
data.head()
# styled_data = data.head().style.format({'ClearRate': "{:.2f}"}).hide().set_table_styles([{"selector":"th.col_heading,td","props":[("width","100px")]}])
# styled_data
#data.values[:5]
#print(data.head().to_string(index=False))

```

## Report Year
Report Year variable is the year of crimes being reported. In this dataset, the data spans from 2014 to 2023, encompassing a ten-year period. No month or date information was given, thus there are only 10 different values for this variable, in a chronological order.

## Category
Category includes information about the nature of the crime. In total there are six crime categroies, they are Crimes Against Property, Crimes Against the Person, Other Federal Statute Violations, Other Criminal Code Violations, Controlled Drugs and Substances Act, and Criminal Code Traffic. They are listed in the table below (@tbl-table2).

```{python}

#| echo: false
#| label: tbl-table2
#| warning: false
#| tbl-cap: Six Crime Categories
import pandas as pd
import plotly.graph_objects as go
data = pd.read_csv('../data/analysis_data/analysis_data.csv')
unique_category = data.Category.unique()
category_df = pd.DataFrame(unique_category, columns=['Categories'])
category_df.index = category_df.index + 1
category_df
```

## Subtype
Under each crime category, there exist multiple subtypes within each category. The following is an exhaustive table (@tbl-table3) of all of the subtypes of crimes and their respective category.

```{python}
#| echo: false
#| label: tbl-table3
#| warning: false
#| tbl-cap: Crime Categories and Subtypes
import pandas as pd
data = pd.read_csv('../data/analysis_data/analysis_data.csv')
unique_category_subtype_data = data.drop_duplicates(subset=['Category', 'Subtype'])
unique_category_subtype_df = unique_category_subtype_data[['Category', 'Subtype']]
unique_category_subtype_df = unique_category_subtype_df.sort_values(by=['Category','Subtype'],ascending=True)
unique_category_subtype_df.reset_index(drop=True, inplace=True)
unique_category_subtype_df.index = unique_category_subtype_df.index + 1
unique_category_subtype_df
```

@@

# Results
## Trend of Reported Cases by Year
During this ten-year-period (2014 - 2023), the overall counts of cases being reported in Toronto has increased significantly, from 113,426 in 2014 to 169,620 in 2023. Meanwhile, a noticable decrease of reported crime counts dropped in 2020 (118,568 cases) and 2021 (120,397 cases), compared to 2019 (144,529 cases). @tbl-table4 below shows the aggregated counts of reported crimes, grouped by their reported years.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table4
#| tbl-cap: Counts of Reported Crimes by Year

import pandas as pd
import matplotlib.pyplot as plt
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

grouped_by_year_df = df.groupby('Year')['Count'].sum().reset_index()
grouped_by_year_df.index = grouped_by_year_df.index +1 
grouped_by_year_df
```

The @fig-figure1 belows depicts the changing pattern of the amounts of reported crimes by year.

```{python}
#| echo: false
#| warning: false
#| label: fig-figure1
#| fig-cap: Trend of Counts of Reported Crimes by Year
import pandas as pd
import matplotlib.pyplot as plt

plt.figure(figsize = (10,6))
plt.plot(grouped_by_year_df['Year'], grouped_by_year_df['Count'], marker = 'o', linestyle = 'solid')
plt.xlabel('Year of The Cases Being Reported')
plt.ylabel('Number of Caese Being Reported')
#plt.title('Count of Reported Cases by Year')
plt.xticks(grouped_by_year_df['Year'])
plt.grid(True)
plt.show()
```

## Trend of Reported Cases per Category by Year
Among all the six categories, crimes against property has a significant higher count than all the other categories. In 2023, a total of 119,629 cases of crimes against property were reported, compared to the second highest count, which is crimes against ther person, with 29,892 cases. Below is a table of the count of reported crime for each category, for 2014, 2020, and 2023, as well as a line chart for the flucuating trend. The @tbl-table5 below shows the amount of reported cases for each category for 2014, 2020, and 2023, respectively.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table5
#| tbl-cap: Trend of Reported Cases per Category in 2014, 2020, and 2023
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

grouped_by_year_category_df = df.groupby(['Year', 'Category'])['Count'].sum().reset_index()
filtered_df = grouped_by_year_category_df[grouped_by_year_category_df['Year'].isin([2014, 2020, 2023])].reset_index()
filtered_df = filtered_df.drop(columns=['index',])
filtered_df.index = filtered_df.index + 1
filtered_df
```

The following is @fig-figure2 showing the trend of how counts of reported crimes varied from 2014 to 2023 for each category.

```{python}
#| echo: false
#| warning: true
#| label: fig-figure2
#| fig-cap: Trend of Reported Cases per Category in 2014, 2020, and 2023
import pandas as pd
import matplotlib.pyplot as plt
plt.figure(figsize=(10, 6))
for category in grouped_by_year_category_df['Category'].unique():
    category_data = grouped_by_year_category_df[grouped_by_year_category_df['Category'] == category]
    plt.plot(category_data['Year'], category_data['Count'], marker='o', linestyle='solid', label=category)
plt.xlabel('Year of The Cases Being Reported')
plt.ylabel('Number of Cases Being Reported')
# plt.title('Count of Reported Cases by Year and Category')
plt.xticks(grouped_by_year_category_df['Year'].unique())
plt.legend(title='Category')
plt.grid(True)
plt.show()
```

## Subtype of Crimes Against Property by Year and Their Gorwth Rates
The overall pattern for each subtype of crimes under crimes against property is comparable to the overall reported crime pattern, with a increase from 2014 to 2023, despite a significant drop in 2020 and 2021. Among all the subtypes within crimes against property, the most common reported crime is theft under $5,000. For each subtype, the nominal growth rate between 2014 and 2023 is also calculated and attached here.
```{python}
#| echo: false
#| warning: false
#| label: tbl-table6
#| tbl-cap: Growth Rate of Each Crime Against Property Subtype from 2014 to 2023
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

crime_against_property_df = df[df['Category'] == 'Crimes Against Property']
#crime_against_property_df.head()
grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year','Subtype'])['Count'].sum().reset_index()
grouped_by_year_subtype_2014_and_2023_df = grouped_by_year_subtype_df[grouped_by_year_subtype_df['Year'].isin([2014,2023])].reset_index().drop(columns=['index',])
grouped_by_year_subtype_2014_and_2023_df.index = grouped_by_year_subtype_2014_and_2023_df.index + 1
#grouped_by_year_subtype_2014_and_2023_df
#This table looks awful. Will pivot it.
pivoted_df = grouped_by_year_subtype_2014_and_2023_df.pivot(index='Subtype', columns='Year', values='Count')
pivoted_df.reset_index(inplace=True)
pivoted_df.columns.name = None 
pivoted_df.columns = ['Subtype', '2014', '2023']
pivoted_df['Growth Rate'] = (((pivoted_df['2023']-pivoted_df['2014'])/pivoted_df['2014'])*100).round(2)
pivoted_df['Growth Rate']=pivoted_df['Growth Rate'].astype(str)+'%'
pivoted_df.index = pivoted_df.index + 1
pivoted_df
```

@fig-figure3 here shows the amount of reported cases for each subtype of crime against property for 2014 and 2023, side by side, respectively.

```{python}
#| echo: false
#| warning: false
#| label: fig-figure3
#| fig-cap: Comparison of Reported Subtypes of Crime Against Property Cases (2014 vs 2023)
import pandas as pd
import matplotlib.pyplot as plt
import numpy as np
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')

crime_against_property_df = df[df['Category'] == 'Crimes Against Property']
#crime_against_property_df.head()
grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year','Subtype'])['Count'].sum().reset_index()
grouped_by_year_subtype_2014_and_2023_df = grouped_by_year_subtype_df[grouped_by_year_subtype_df['Year'].isin([2014,2023])].reset_index().drop(columns=['index',])
grouped_by_year_subtype_2014_and_2023_df.index = grouped_by_year_subtype_2014_and_2023_df.index + 1
#grouped_by_year_subtype_2014_and_2023_df
#This table looks awful. Will pivot it.
pivoted_df = grouped_by_year_subtype_2014_and_2023_df.pivot(index='Subtype', columns='Year', values='Count')
pivoted_df.reset_index(inplace=True)
pivoted_df.columns.name = None 
pivoted_df.columns = ['Subtype', '2014', '2023']
pivoted_df['Growth Rate'] = (((pivoted_df['2023']-pivoted_df['2014'])/pivoted_df['2014'])*100).round(2)
pivoted_df['Growth Rate']=pivoted_df['Growth Rate'].astype(str)+'%'
pivoted_df.index = pivoted_df.index + 1

subtypes = pivoted_df['Subtype']
counts_2014 = pivoted_df['2014']
counts_2023 = pivoted_df['2023']
fig, ax = plt.subplots(figsize=(12, 6))
bar_width = 0.35
index = np.arange(len(subtypes))
bars_2014 = ax.bar(index, counts_2014, bar_width, label='Count in 2014', color='blue')
bars_2023 = ax.bar(index + bar_width, counts_2023, bar_width, label='Count in 2023', color='green')
ax.set_xlabel('Subtype')
ax.set_ylabel('Count of Cases')
ax.set_xticks(index + bar_width / 2)
ax.set_xticklabels(subtypes, rotation=45, ha='right')
ax.legend()
plt.tight_layout()
plt.grid(True)
plt.show()
```

## Crime Subtype Clear Rate from 2014 to 2023 and Its Trend
Crime clearing rate has dropped for all subtypes of crimes against property from 2014 to 2023, ranging from a 10.25% decrease for break and enter to apartment, to 87.02% decrease for fraud. The clearing rate for 2014 and 2023, as well as the change in clear rate for each sutype of crime against property are listed below in @tbl-table7.

```{python}
#| echo: false
#| warning: false
#| label: tbl-table7
#| fig-cap: Case Clear Rate and Change from 2014 to 2023 for Crime Against Property Subtypes
import pandas as pd
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')
crime_against_property_df = df[df['Category'] == 'Crimes Against Property']
grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year', 'Subtype'])['ClearRate'].mean().reset_index()
pivoted_clear_rate_df = grouped_by_year_subtype_df.pivot(index='Subtype', columns='Year', values='ClearRate')
pivoted_clear_rate_df = pivoted_clear_rate_df[[2014, 2023]]
pivoted_clear_rate_df['Clear Rate Change'] = (((pivoted_clear_rate_df[2023]-pivoted_clear_rate_df[2014])/pivoted_clear_rate_df[2014])*100).round(2)
pivoted_clear_rate_df['Clear Rate Change']=pivoted_clear_rate_df['Clear Rate Change'].astype(str)+'%'
pivoted_clear_rate_df.columns.name = None
pivoted_clear_rate_df.reset_index(inplace=True)
pivoted_clear_rate_df.index=pivoted_clear_rate_df.index +1
pivoted_clear_rate_df

```

@fig-figure4 show the trend of the change in case clearing rate of each subtype of crime against property from 2014 to 2023.

```{python}
#| echo: false
#| warning: false
#| label: fig-figure4
#| fig-cap: Trends of Case Clear Rate Change from 2014 to 2023 for Crime Against Property Subtypes
import pandas as pd
import matplotlib.pyplot as plt
df = pd.read_csv('/Users/jiazhoubi/Desktop/UT MI/INF 3104/reported-crime/data/analysis_data/analysis_data.csv')
crime_against_property_df = df[df['Category'] == 'Crimes Against Property']

grouped_by_year_subtype_df = crime_against_property_df.groupby(['Year','Subtype'])['ClearRate'].mean().reset_index()
plt.figure(figsize=(10,6))
for subtype in grouped_by_year_subtype_df['Subtype'].unique():
    subtype_data = grouped_by_year_subtype_df[grouped_by_year_subtype_df['Subtype'] == subtype]
    plt.plot(subtype_data['Year'],subtype_data['ClearRate'],marker = 'o', linestyle = 'solid', label=subtype)
plt.xlabel('Year of The Cases Being Reported')
plt.ylabel('Clearing Rate of Cases Being Reported')
plt.xticks(grouped_by_year_subtype_df['Year'].unique())
plt.legend(title='Subtype',loc = 'upper left',fontsize = 8)
plt.grid(True)
plt.show()
```
